---
title: "Student Performance Dataset Analysis"
author: "Group 14- Marissa Ashner, Ben Bulen, Kyle Grosser, Benjana Guraziu, Kushal Shah"
date: "4/24/2019"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
h1.title {
  font-size: 24px;
}
code.r{ /* Code block */
    font-size: 10px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 10px;
}
</style>

```{r, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

```{r, warning = FALSE}
student <- read.table("student-mat.csv", sep=";", header=T)
student <- as_tibble(student)
```

## Data Description 

The inital covariates in the dataset are:

1. SCHOOL - student's school (binary: "GP" - Gabriel Pereira or "MS" - Mousinho da Silveira)
2. SEX - student's sex (binary: "F" - female or "M" - male)
3. AGE - student's age (numeric: from 15 to 22)
4. ADDRESS - student's home address type (binary: "U" - urban or "R" - rural)
5. FAMSIZE - family size (binary: "LE3" - less or equal to 3 or "GT3" - greater than 3)
6. PSTATUS - parent's cohabitation status (binary: "T" - living together or "A" - apart)
7. MEDU - mother's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8. FEDU - father's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9. MJOB - mother's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
10. FJOB - father's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
11. REASON - reason to choose this school (nominal: close to "home", school "reputation", "course" preference or "other")
12. GUARDIAN - student's guardian (nominal: "mother", "father" or "other")
13. TRAVELTIME - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14. STUDYTIME - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15. FAILURES - number of past class failures (numeric: n if 1<=n<3, else 4)
16. SCHOOLSUP - extra educational support (binary: yes or no)
17. FAMSUP - family educational support (binary: yes or no)
18. PAID - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19. ACTIVITIES - extra-curricular activities (binary: yes or no)
20. NURSERY - attended nursery school (binary: yes or no)
21. HIGHER - wants to take higher education (binary: yes or no)
22. INTERNET - Internet access at home (binary: yes or no)
23. ROMANTIC - with a romantic relationship (binary: yes or no)
24. FAMREL - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25. FREETIME - free time after school (numeric: from 1 - very low to 5 - very high)
26. GOOUT - going out with friends (numeric: from 1 - very low to 5 - very high)
27. DALC - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28. WALC - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29. HEALTH - current health status (numeric: from 1 - very bad to 5 - very good)
30. ABSENCES - number of school absences (numeric: from 0 to 93)

And the response variables are: 

1. G1 - first period grade (numeric: from 0 to 20) 
2. G2 - second period grade (numeric: from 0 to 20) 
3. G3 - final grade (numeric: from 0 to 20, output target)

For this analysis, we will be considering the `G3` final grade as our response. 

## Expoloratory Data Analysis and Variable Filtering

### Distribution of the Response Variable 

```{r}
student %>% 
  ggplot(aes(G3)) + geom_histogram(bins = 20)
```

```{r}
student %>% select(G1, G2, G3) %>% filter(G3 ==0 & G1 ==0 & G2 ==0)
```

Let's remove the students with a grade of 0, as that doesn't make much practical sense, especially since all of these students have grades higher than 0 first and/or second period.

```{r}
student <- student %>% filter(G3 != 0) %>% select(-G1, -G2)
student %>% 
  ggplot(aes(G3)) + geom_histogram(bins = 20)
```


### Covariates and their relation to Final Grade

```{r}
student %>% 
  gather(key = "Variable", value = "Value", -G3) %>% 
  ggplot(aes(Value, G3)) + facet_wrap(~ Variable, scales = "free") + 
  geom_boxplot()
```

### Correlated Variable Filtering

### Covariates of Potential Interest

## Models and Model Testing

## Model Selection and Interpretation

## Conclusion

