---
title: "Student Performance Dataset Analysis"
author: "Group 14- Marissa Ashner, Ben Bulen, Kyle Grosser, Benjana Guraziu, Kushal Shah"
date: "4/24/2019"
output: html_document
---

<style type="text/css">

body{ /* Normal  */
      font-size: 12px;
  }
h1.title {
  font-size: 24px;
}
code.r{ /* Code block */
    font-size: 10px;
}
pre { /* Code block - determines code spacing between lines */
    font-size: 10px;
}
</style>

```{r, include = FALSE, warning = FALSE}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
library(tidyverse)
```

```{r, warning = FALSE}
student <- read.table("student-mat.csv", sep=";", header=T)
student <- as_tibble(student)
```

## Data Description 

The inital covariates in the dataset are:

1. SCHOOL - student's school (binary: "GP" - Gabriel Pereira or "MS" - Mousinho da Silveira)
2. SEX - student's sex (binary: "F" - female or "M" - male)
3. AGE - student's age (numeric: from 15 to 22)
4. ADDRESS - student's home address type (binary: "U" - urban or "R" - rural)
5. FAMSIZE - family size (binary: "LE3" - less or equal to 3 or "GT3" - greater than 3)
6. PSTATUS - parent's cohabitation status (binary: "T" - living together or "A" - apart)
7. MEDU - mother's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
8. FEDU - father's education (numeric: 0 - none,  1 - primary education (4th grade), 2 – 5th to 9th grade, 3 – secondary education or 4 – higher education)
9. MJOB - mother's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
10. FJOB - father's job (nominal: "teacher", "health" care related, civil "services" (e.g. administrative or police), "at_home" or "other")
11. REASON - reason to choose this school (nominal: close to "home", school "reputation", "course" preference or "other")
12. GUARDIAN - student's guardian (nominal: "mother", "father" or "other")
13. TRAVELTIME - home to school travel time (numeric: 1 - <15 min., 2 - 15 to 30 min., 3 - 30 min. to 1 hour, or 4 - >1 hour)
14. STUDYTIME - weekly study time (numeric: 1 - <2 hours, 2 - 2 to 5 hours, 3 - 5 to 10 hours, or 4 - >10 hours)
15. FAILURES - number of past class failures (numeric: n if 1<=n<3, else 4)
16. SCHOOLSUP - extra educational support (binary: yes or no)
17. FAMSUP - family educational support (binary: yes or no)
18. PAID - extra paid classes within the course subject (Math or Portuguese) (binary: yes or no)
19. ACTIVITIES - extra-curricular activities (binary: yes or no)
20. NURSERY - attended nursery school (binary: yes or no)
21. HIGHER - wants to take higher education (binary: yes or no)
22. INTERNET - Internet access at home (binary: yes or no)
23. ROMANTIC - with a romantic relationship (binary: yes or no)
24. FAMREL - quality of family relationships (numeric: from 1 - very bad to 5 - excellent)
25. FREETIME - free time after school (numeric: from 1 - very low to 5 - very high)
26. GOOUT - going out with friends (numeric: from 1 - very low to 5 - very high)
27. DALC - workday alcohol consumption (numeric: from 1 - very low to 5 - very high)
28. WALC - weekend alcohol consumption (numeric: from 1 - very low to 5 - very high)
29. HEALTH - current health status (numeric: from 1 - very bad to 5 - very good)
30. ABSENCES - number of school absences (numeric: from 0 to 93)

And the response variables are: 

1. G1 - first period grade (numeric: from 0 to 20) 
2. G2 - second period grade (numeric: from 0 to 20) 
3. G3 - final grade (numeric: from 0 to 20, output target)

For this analysis, we will be considering the `G3` final grade as our response. 

## Expoloratory Data Analysis and Variable Filtering

### Distribution of the Response Variable 

```{r}
student %>% 
  ggplot(aes(G3)) + geom_histogram(binwidth = 1, color='black', fill = 'grey') + 
  geom_vline(xintercept = mean(student$G3), lwd = 1, color = 'red') +
  labs(title = "Distribution of Final Math Grade", x="Final Grade", y="Count")
```

```{r}
student %>% select(G1, G2, G3) %>% filter(G3 ==0 & G1 ==0 & G2 ==0)
```

Let's remove the students with a grade of 0, as that doesn't make much practical sense, especially since all of these students have grades higher than 0 first and/or second period.
**BB: agree with this approach, if we are feeling adventurous we could treat those 0's as missing and impute them but let's not over complicate things yet

```{r}
student <- student %>% filter(G3 != 0) %>% select(-G1, -G2)
student %>% 
  ggplot(aes(G3)) + geom_histogram(binwidth = 1, color='black', fill = 'grey') + 
  geom_vline(xintercept = mean(student$G3), lwd = 1, color = 'red') +
  labs(title = "Distribution of Final Math Grade", x="Final Grade", y="Count")
```

### Covariates and their relation to Final Grade

```{r}
student %>% 
  gather(key = "Variable", value = "Value", -G3) %>% 
  ggplot(aes(Value, G3)) + facet_wrap(~ Variable, scales = "free") + 
  geom_boxplot()
```

### Collinearity Analysis 

VIF and Tolerance 

```{r}
covariates_int <- model.matrix(G3 ~ ., data = student )
covariates <- covariates_int %>% as.data.frame() %>% select(-`(Intercept)`)
student_d <- data.frame(G3 = student$G3, covariates)
model1 <- lm(G3 ~ ., data = student_d)
VIF = car::vif(model1)
Tol = 1/VIF
df = data.frame(Variable = names(VIF), VIF = VIF, Tolerance = Tol)
df %>% arrange(desc(VIF)) %>% head(10) %>% knitr::kable(align = c("c", "c"),)
```

Displayed above are the covariates with the highest Variance Inflation Factors.

EigenAnalysis for the Correlation and Scaled SSCP Matrices 

```{r}
xtx = t(covariates_int)%*%covariates_int
Ds_half <- diag(diag(xtx)^-0.5)
sscp <- Ds_half %*% xtx %*% Ds_half
eig_sscp <- eigen(sscp)$values
PCs_sscp <- prcomp(sscp)[2]
CI_sscp <- sqrt(eig_sscp[1]/eig_sscp)

cov_center <- apply(covariates, 2, function(y) y - mean(y))
C <- (t(cov_center)%*%cov_center)/dim(cov_center)[1]
Dc_half <- diag(diag(C)^-0.5)
R <- Dc_half %*% C %*% Dc_half
eig_corr <- eigen(R)$values
CI_corr <- sqrt(eig_corr[1]/eig_corr)
PCs_corr <- prcomp(R)[2]

df <- data.frame("Eigenvalue" = c(eig_corr), "Condition Index" = c(CI_corr))
df2 <- data.frame("Eigenvalue" = c(eig_sscp), "Condition Index" = c(CI_sscp))

df %>% tail(2) %>% knitr::kable(align = c("c", "c"))
df2 %>% filter(`Condition.Index` > 30) %>% knitr::kable(align = c("c", "c"))
```

From the Correlation Matrix, the highest Condition Index is 7.0754, which means no collinearity needs to be taken care of there. From the Scaled SSCP Matrix, there are two Condition Indices over 30.

```{r}
PCs_sscp$rotation[1:17,39:40]
```

Above shows the last 2 PC's of the Scaled SSCP Matrix for the 1st-17th covariates (the rest have values close to 0). We can see that the 8th, 15th, and 16th covariates may be collinear with the intercept. These covariates are listed below.

```{r}
names(VIF)[c(4, 8, 15, 16)]
```

** HOW TO PICK WHAT TO REMOVE ** should we just take out the intercept?

### Covariates of Potential Interest

## Models and Model Testing

### Model 1: Full Model 
*BB: since categorical vars are being grouped into several vars we have more than 30 Betas. I count 39 below.

$$y = \beta_0 + \beta_1X_1 + ... + \beta_{30}X_{30} + \epsilon$$

```{r}
summary(model1)
```

** ADJUSTED R2 of 0.256 BOO ** 

#### Diagnostics

```{r}
stud_resid = rstudent(model1)
dat_resid <- data.frame(RESID = stud_resid, ABS = abs(stud_resid), PRED = model1$fitted.values)
dat_resid %>% arrange(-ABS) %>% filter(ABS > 2)
```

Since this set of largest studentized residuals are all greater than 2, it may be of interest to examine them in further detail.

```{r}
nortest::ad.test(dat_resid$RESID)
```

From the code above, since the p-value 0.4184 is greater than $\alpha = 0.05$, we accept the null hypothesis that the residuals are normally distributed. 

```{r, fig.height=4, fig.width=8}
p1 <- ggplot(data = dat_resid) + geom_histogram(aes(RESID), bins = 25) + labs(x = "Studentized Residuals")
p2 <- ggplot(data = dat_resid) + geom_point(aes(PRED, RESID)) + geom_hline(aes(yintercept = 0)) + labs(x = "Predicted Values", y = "Studentized Residuals")
cowplot::plot_grid(p1, p2)
```

### Model 2: No Intercept

```{r}
model2 <- lm(G3 ~ . -1, data = student_d)
summary(model2)
```

** ADJUSTED R2 OF .9421 !!!!!! ** 

#### Diagnostics

```{r}
stud_resid = rstudent(model2)
dat_resid <- data.frame(RESID = stud_resid, ABS = abs(stud_resid), PRED = model1$fitted.values)
dat_resid %>% arrange(-ABS) %>% filter(ABS > 2)
```

```{r}
nortest::ad.test(dat_resid$RESID)
```

```{r, fig.height=4, fig.width=8}
p1 <- ggplot(data = dat_resid) + geom_histogram(aes(RESID), bins = 25) + labs(x = "Studentized Residuals")
p2 <- ggplot(data = dat_resid) + geom_point(aes(PRED, RESID)) + geom_hline(aes(yintercept = 0)) + labs(x = "Predicted Values", y = "Studentized Residuals")
cowplot::plot_grid(p1, p2)
```

## Model Selection and Interpretation

## Conclusion

